Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
Make checking consistent with new terminology & add consistency clause for 'both'

\section{Inference and checking}
\label{sec:analysis}

@@

Verification occurs when a user supplied specification is compared to the
actual access pattern. The mathematical model developed in
Section~\ref{sec:semantics} is used this comparison.

A specification is first synthesised into its model. Then access patterns
are checked against the model for consistency as defined in
Figure~\ref{fig:consistency}. The consistency predicate is defined recursively.
It first checks multiplicity if $\textsf{once}$ is at the top level, \ie{}
specification in question involves \texttt{readOnce} modifier, then array
accesses are checked for uniqueness using $unique$ predicate and the problem
is reduced to $\textsf{mult}$, which does not impose any extra constraints.
Next, consistency predicate considers whether the stencil is an approximation.
Semantically, a lower bound is achieved if every unifier in the model
correspond to some relativised vector of indices as that means all cells
described in the specification are part of the access pattern. Conversely,
an upper bound means all observed access patterns have a unifier. Hence,
the actual cells are a subset of the specified ones. In combination an
exact specification, \ie{} one without any approximation modifiers, is
one that is both a lower and an upper bound.

@@

  The predicate is reduced to
  $cons(\textsf{once}(\textsf{exact}(m)),L3,b)$.
%
\item $cons$ predicate from the previous step evaluates to
  $$unique(L3,b) \; \wedge \; cons(\textsf{mult}(\textsf{exact}(m)),L3,b).$$
%
  The former condition holds as no access pattern is repeated in $L3$ by simple
  inspection.
%
\item We further expand the function to 
  \begin{equation*}
    cons(\textsf{mult}(\textsf{up}(m)),L3,b) \wedge 
    cons(\textsf{mult}(\textsf{low}(m),L3,b).
  \end{equation*}
%
\item To evaluate the two predicates above, we need relativised indices for the
  array in question. This is obtained by $analyse$ function from
  Subsection~\ref{subsec:inference}.

  For the example above the relativised indices are as follows:
%
\begin{dmath*}
analyse(L3,b) =
  \{ \vtwo{0}{-1},
     \vtwo{0}{0},
     \vtwo{0}{1},
     \vtwo{-1}{0},
     \vtwo{1}{0}
  \}
\end{dmath*}
%
\item Below we present two match ups of relativised access patterns and unifiers
  to show both the lower and the upper bound holds.

  The unifications for the upper bound:
  \begin{align*}
     \vtwoh{0}{-1} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{0}{1} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{-1}{0} & \sim_2 \vtwoh{\infty}{0} \\
     \vtwoh{1}{0} & \sim_2 \vtwoh{\infty}{0}
  \end{align*}

  The unifications for the lower bound:
  \begin{align*}
     \vtwoh{-1}{0} & \sim_2 \vtwoh{-1}{\infty} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{1}{0} & \sim_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1} & \sim_2 \vtwoh{\infty}{-1} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{\infty}{0} \\
     \vtwoh{0}{1} & \sim_2 \vtwoh{\infty}{1}
  \end{align*}

  Since both predicates are satisfied the array accesses specified on line three
  are consistent with the \texttt{fivepoint} specification.
\end{enumerate}

Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j)}, the
new assignment $L3'$ becomes:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
The first four steps will proceed the same way and in the fifth step result
of $\mathit{analyse}(L3',b)$ will have the extra term $\vtwoh{2}{2}$. In sixth
step, since the model has not changed and the relativised indices are only
augmented with a new element, the unification for lower bound still hold.
However, there are not elements in the model, $m$, that can unify with
$\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.

Note that if the specification was as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b}
\end{minted}
%
The consistency function would have held because for the upper bound
$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would never have
been needed. This semantically makes sense since the region
$\texttt{fivepoint}$ is indeed a lower bound for the access pattern of
$b$.

>>>>

\section{Inference, checking, and synthesis}
\label{sec:analysis}

@@

In checking mode, the specification is verified with respect to the indexing
scheme of the actual access pattern. The mathematical model of the specification
developed in Section~\ref{sec:semantics} is used this process.

A specification is first synthesised into its model. Then occurrences of an
array in an assignment are checked against the model for consistency as defined
in Figure~\ref{fig:consistency}. The consistency predicate is defined
recursively.

It first checks multiplicity if $\textsf{once}$ is at the top
level, \ie{} specification in question involves \texttt{readOnce} modifier, then
the array accesses are checked for uniqueness using $unique$ predicate and the
problem is reduced to $\textsf{mult}$, which does not impose any extra
constraints.

Next, consistency predicate considers whether the stencil is an approximation.
Semantically, a lower bound is achieved if every indexing scheme in the
model generalises some relativised access pattern that means all cells
described in the specification are part of the access pattern. Conversely,
an upper bound means all observed access patterns can be generalised with
the indexing schemes of the model. Hence, the actual cells are a subset of the
specified ones. In combination an exact specification, \ie{} one without any
approximation modifiers, is one that is both a lower and an upper bound.

@@

  The predicate is reduced to
  $cons(\textsf{once}(\textsf{exact}(m)),L3,b)$.
%
\item $cons$ predicate from the previous step evaluates to
  $$unique(L3,b) \; \wedge \; cons(\textsf{mult}(\textsf{exact}(m)),L3,b).$$
%
  The former condition holds as no access pattern is repeated in $L3$ by simple
  inspection.
%
\item We further expand the function to 
  \begin{equation*}
    cons(\textsf{mult}(\textsf{up}(m)),L3,b) \wedge 
    cons(\textsf{mult}(\textsf{low}(m),L3,b).
  \end{equation*}
%
\item To evaluate the two predicates above, we need indexing schemes for the
  array in question. This is obtained by $analyse$ function from
  Subsection~\ref{subsec:inference}.

  For the example above the indexing schemes are as follows:
%
\begin{dmath*}
analyse(L3,b) =
  \{ \vtwo{0}{-1},
     \vtwo{0}{0},
     \vtwo{0}{1},
     \vtwo{-1}{0},
     \vtwo{1}{0}
  \}
\end{dmath*}
%
\item Below we present two sets of generalisations to show both the lower and
  the upper bound holds.

  The generalisations for the upper bound:
  \begin{align*}
     \vtwoh{0}{-1} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{0}{1} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{-1}{0} & \sim_2 \vtwoh{\infty}{0} \\
     \vtwoh{1}{0} & \sim_2 \vtwoh{\infty}{0}
  \end{align*}

  The generalisations for the lower bound:
  \begin{align*}
     \vtwoh{-1}{0} & \sim_2 \vtwoh{-1}{\infty} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{1}{0} & \sim_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1} & \sim_2 \vtwoh{\infty}{-1} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{\infty}{0} \\
     \vtwoh{0}{1} & \sim_2 \vtwoh{\infty}{1}
  \end{align*}

  Since both predicates are satisfied the array accesses specified on line
  three, the stencil comutation is consistent with the \texttt{fivepoint}
  specification.
\end{enumerate}

Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j+2)},
the new assignment $L3'$ becomes:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
The first four steps will proceed the same way. In the fifth step, result
of $\mathit{analyse}(L3',b)$ will have the extra term $\vtwoh{2}{2}$. In the
sixth step, since the model has not changed and the indexing schemes are only
augmented with a new element, the generalisations for lower bound still hold.
However, there are no elements in the model, $m$, that can generalise
$\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.

Note that if the specification was as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b
\end{minted}
%
The consistency predicate would have held because for the upper bound
$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would not have
been needed. This semantically makes sense since the region
$\texttt{fivepoint}$ is indeed contained in the regions touched by $b$.

\subsection{Synthesis\label{subsec:synthesis}}
